---
title: "Introduction to AEOLUS dataset"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F)
```

## Brief Introduction

```{r load}

library(fst)
library(data.table)
library(tidyverse)

theme_set(theme_classic())

data_dir <- "../data/"
aeolus <- read_fst(paste0(data_dir,"aeolus.fst"),as.data.table = T)

```

## Brief Introduction

The Adverse Event Open Learning through Universal Standardization, or AEOLUS, is detailed in the paper by [Banda et al](http://www.nature.com/articles/sdata201626).

Briefly, it contains a standardized version of reports from the FDA's Adverse Event Reporting System, allowing for efficient and reproducible analyses of ADEs. 

The aeolus dataset has `r scales::comma(nrow(aeolus))` rows with `r scales::comma(length(unique(aeolus[,id])))` unique reports.

Below, I briefly show characteristics of this dataset. 

## Table display

```{r display}

aeolus[1:10]

```

## Number of drugs reported per report

```{r num_drugs_per_report}

unique(aeolus[,.(id,drug_concept_id)])[
  ,.N,by=id
][
  order(-N)
] %>% 
  ggplot(aes(N)) +
  geom_histogram() +
  scale_x_log10() +
  theme(
    text = element_text(face="bold",size=16)
  ) +
  ylab("Number of Reports") +
  xlab("Number of Drugs")

```

## Number of outcomes reported per report

```{r num_outcomes_per_report}

unique(aeolus[,.(id,outcome_concept_id)])[
  ,.N,by=id
][
  order(-N)
] %>% 
  ggplot(aes(N)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_continuous(labels = scales::comma) + 
  xlab("Number of Reports") +
  ylab("Number of Outcomes") +
  theme(
    text = element_text(face="bold",size=16)
  )

```

## Number of ADRs reported 

```{r num_adrs_per_report}

unique(aeolus[,.(id,drug_concept_id,outcome_concept_id)])[
  ,.N,by=id
][
  order(-N)
] %>% 
  ggplot(aes(N)) +
  geom_histogram() +
  scale_x_log10() +
  scale_y_continuous(labels = scales::comma) + 
  xlab("Number of Reports") +
  ylab("Number of ADRs") +
  theme(
    text = element_text(face="bold",size=16)
  )

```

## Number of reports per year

```{r reports_per_year}

unique(aeolus[,.(id,report_year)])[
  ,.N,by=report_year
][
  order(report_year)
][
  ,.(report_year = as.character(report_year),
     N)
] %>% 
  ggplot(aes(report_year,N,fill=report_year)) +
  geom_bar(stat="identity") +
  scale_y_continuous(labels = scales::comma) + 
  xlab("Report Year") +
  ylab("Number of Reports") +
  theme(
    axis.text.x = element_text(angle=45,vjust=0.7),
    text = element_text(face="bold",size=16),
    legend.position = "none"
  )

```

## Number of reports per age category

```{r reports_per_age_cat}

data <- na.omit(
  unique(
    aeolus[,.(id,dev_age_cat)]
    )[
  ,.N,by=dev_age_cat
  ]
)[
  order(dev_age_cat)
][
  ,.(dev_age_cat = as.character(dev_age_cat),
     N)
] 

age_cats <- as.character(unique(data[,dev_age_cat]))
ord <- order(sapply(strsplit(age_cats,"[\\[(]"),function(x){strsplit(x[2],",")[[1]][1]}) %>% as.integer())
seq_age_cats <- age_cats[ord]

data$dev_age_cat <- factor(data$dev_age_cat,levels=seq_age_cats)


data %>% 
  ggplot(aes(dev_age_cat,N,fill=dev_age_cat)) +
  geom_bar(stat="identity") +
  scale_y_continuous(trans="log10",labels = scales::comma) + 
  xlab("Human Development Age Category") +
  ylab("Number of Reports") +
  theme(
    axis.text.x = element_text(angle=45,vjust=0.7),
    text = element_text(face="bold",size=16),
    legend.position = "none"
  )

data <- na.omit(
  unique(
    aeolus[,.(id,age_cat)]
    )[
  ,.N,by=age_cat
  ]
)[
  ,.(age_cat = as.character(age_cat),
     N)
]

age_cats <- as.character(unique(data[,age_cat]))
ord <- order(sapply(strsplit(age_cats,"[\\[(]"),function(x){strsplit(x[2],",")[[1]][1]}) %>% as.integer())
seq_age_cats <- age_cats[ord]

data$age_cat <- factor(data$age_cat,levels=seq_age_cats)

data %>% 
  ggplot(aes(age_cat,N,fill=age_cat)) +
  geom_bar(stat="identity") +
  scale_y_continuous(labels = scales::comma) + 
  xlab("5-year Age Category") +
  ylab("Number of Reports") +
  theme(
    axis.text.x = element_text(angle=45,vjust=0.7),
    text = element_text(face="bold",size=16),
    legend.position = "none"
  )

```

## Condition types

```{r condition_types}

aeolus[
  ,.(id,outcome_domain_id)
  ][
    ,.N,
    by=outcome_domain_id
  ][
    order(-N)
  ] %>% 
  ggplot(aes(outcome_domain_id,N,fill=outcome_domain_id)) +
  geom_bar(stat="identity") +
  scale_y_continuous(trans="log10",labels=scales::comma) +
  coord_flip() +
  ylab("Number of Reports") +
  xlab("Condition domain") +
  theme(
    legend.position = "none",
    text = element_text(face="bold",size=16)
  )
  
```

## ATC 1st distribution of drugs

```{r atc_1,fig.height=10}

df <- aeolus[!is.na(atc_1st),
       .(id,atc_1st)][
         ,.N,
         by=atc_1st
       ][
         order(-N)
       ]

df$atc_1st <- factor(df$atc_1st,levels=unique(df$atc_1st))

df %>% 
  ggplot(aes(atc_1st,N,fill=atc_1st)) +
  geom_bar(stat="identity") +
  scale_y_continuous(trans="log10",labels=scales::comma) +
  coord_flip() +
  xlab("System of drug action") +
  ylab("Number of Reports") + 
  guides(
    fill=guide_legend(
      ncol = 1,
      title.position = "top")
    ) +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size=0),
    text = element_text(face="bold",size=10)
  )

```