---
title: "Exploratory Data Analysis of young age category biased adverse drug reactions"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F,warning = F)
```

## Introduction

This notebook serves as the initialv exploration of the calculated pharmacovigilance statistics for determining ADRs disproportionally within young age categories. The question of whether these ADRs would be associated to child development (where age associations are used here as a proxy for) is yet to be determined. This analysis serves as the initial step in pursuit of developmentally-regulated ADRs. 

## Set up

```{r setvars}
library(tidyverse)
theme_set(theme_classic(base_size = 16,base_family = "Times New Roman"))
library(data.table)

data_dir <- "../data/"

type <- c("ADR","ADR_class","hADR")
dir_name <- paste0(data_dir,"ADRs_PhV/ATC/ATC_3rd/ANTIEPILEPTICS/ALL_DRUGS/")

files <- list.files(dir_name,"*annotated.csv")

phv <- lapply(
  files,
  function(x){
    fread(paste0(dir_name,x))
  })
names(phv) <- c("ADR","ADR_class","hADR")



phv_binded <- bind_rows(phv,.id="type")

dev_age_cats <- as.character(unique(phv_binded$dev_age_cat))
ord <- order(sapply(strsplit(dev_age_cats,"[\\[(]"),function(x){strsplit(x[2],",")[[1]][1]}) %>% as.integer())
phv_binded$dev_age_cat <- factor(phv_binded$dev_age_cat,levels=dev_age_cats[ord])

```

## Number of ADR candidates

```{r Number_ADR_candidates_per_type}

candidates <- unique(
  phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,type)]
  )

data <- candidates[,.N,by=type]

data %>% 
  ggplot(aes(type,N,fill=type)) + 
  geom_bar(stat="identity") + 
  geom_text(aes(label=N,y=N+5)) +
  scale_fill_brewer(palette="Set1") +
  ylab("Number of ADRs") + 
  xlab("") +
  theme(
    text = element_text(face="bold",size=16)
  )
```



## ADRs by ADR type

```{r frequent_candidate_ADRs_per_type,fig.width=14,fig.height=10}

candidates <- unique(
  phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,type)]
  )

data <- candidates[,.N,by=.(outcome_concept_name,type)][order(-N)]

data %>%
  group_by(type) %>% 
  top_n(5) %>% 
  ungroup() %>% 
  ggplot(aes(reorder(outcome_concept_name,N),N,fill=outcome_concept_name)) + 
  geom_bar(stat="identity") + 
  facet_grid(type~.,scales="free") +
  ylab("N") +
  xlab("ADRs") +
  coord_flip() +
  guides(
    fill = guide_legend(title = "ADR",
                        title.position = "top",
                        label.position = "left",
                        ncol=3)
    ) +
  theme(
    legend.text = element_text(size=10,face="bold"),
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )

```

## Drugs by ADR type

```{r Drug_frequency_in_candidate_ADRs_per_type,fig.height=7}

candidates <- unique(
  phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,type)]
  )

data <- candidates[,.N,by=.(drug_concept_name,type)][order(-N)]

data %>%
  ggplot(aes(reorder(drug_concept_name,N),
             N,fill=drug_concept_name)) + 
  geom_bar(stat="identity") + 
  facet_grid(type~.,scales="free") +
  ylab("N") +
  xlab("Drugs") +
  coord_flip() +
  #scale_fill_brewer(palette="Set1") +
  scale_y_continuous(labels=scales::number) + 
  guides(
    fill = guide_legend(title = "",
                        label.position = "left",
                        ncol=3)
    ) +
  theme(
    legend.text = element_text(size=16,face="bold"),
    text = element_text(face="bold",size=10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_text(size=16),
    legend.position = "bottom"
    )

```

## Number of ADRs per type by age category

```{r Number_of_candidate_ADRs_per_type_by_dev_age_cat,fig.width=10,fig.height=6}

candidates <- phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,dev_age_cat,
               type,PRR_mse,PRR_norm)][
                 PRR_mse>2&PRR_norm>2][
                 ,.(drug_concept_name,outcome_concept_name,
                    dev_age_cat,type)] %>% 
  unique()

data <- candidates[,
                   .N,
                   by=.(dev_age_cat,type)][order(-N)]

dev_age_cats <- as.character(unique(data$dev_age_cat))
ord <- order(sapply(strsplit(dev_age_cats,"[\\[(]"),function(x){strsplit(x[2],",")[[1]][1]}) %>% as.integer())
data$dev_age_cat <- factor(data$dev_age_cat,levels=dev_age_cats[ord])

data %>% 
  ggplot(aes(dev_age_cat,N,fill=type)) +
  geom_bar(stat="identity",position = "dodge") +
  scale_fill_brewer(palette="Set1") +
  theme(
    text = element_text(face="bold")
  )
  

```

## PRR_norm distribution of ADRs per type and dev_age_cat

```{r PRR_norm_distribution_per_ADR_type_and_dev_age_cat,fig.width=10,fig.height=10}

candidates <- unique(
  phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,dev_age_cat,PRR_norm,type)]
  )

data <- candidates[,.N,by=.(outcome_concept_name,type)][order(-N)]

candidates %>% 
  ggplot(aes(PRR_norm,fill=type)) + 
  geom_histogram() + 
  scale_x_log10() +
  facet_grid(dev_age_cat~type) +
  scale_fill_brewer(palette="Set1") +
  theme(
    text = element_text(face="bold",size=16)
  )
```

## Top ADRs per type in age categories

```{r top_candidate_ADRs_per_type_in_dev_age_cat,fig.width=30,fig.height=30}

candidates <- phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,dev_age_cat,
               type,PRR_mse,PRR_norm)][
                 PRR_mse>2&PRR_norm>2]

candidates %>% 
  group_by(drug_concept_name,outcome_concept_name,dev_age_cat,PRR_norm) %>% 
  top_n(5,PRR_norm) %>% 
  ungroup() %>% 
  ggplot(aes(outcome_concept_name,PRR_norm,fill=drug_concept_name)) +
  geom_bar(stat="identity",position = "dodge") +
  facet_grid(drug_concept_name~dev_age_cat,scales = "free_y") +
  coord_flip() +
  xlab("") +
  ylab("PRR_norm") +
  theme_bw() +
  theme(
    legend.position = " bottom",
    text = element_text(face="bold",size=24),
    axis.text.x = element_text(size=20),
    axis.text.y = element_text(size=14)
  )
  

```
## Number of ADRs shared between dev_age_cats

```{r Number_of_shared_ADRs_per_type_between_dev_age_cat}

candidates <- phv_binded[candidate==T,
             .(drug_concept_name,outcome_concept_name,dev_age_cat,
               type,PRR_mse,PRR_norm)][
                 PRR_mse>2&PRR_norm>2]

data <- candidates[
  ,.N,
  by=.(drug_concept_name,outcome_concept_name)][
    ,.(shared = .N),
    by=N
    ]

data$N <- factor(data$N)

data %>% 
  ggplot(aes(N,shared,fill=N)) +
  geom_bar(stat="identity") +
  ylab("Number of ADRs\nshared across age categories") +
  xlab("Number of shared ADRs") +
  theme(
    legend.position = "none",
    text = element_text(face="bold",size=24)
  )
```